---
title: "Semester / Summer Report"
format:
  typst:
    linkcolor: "#0066cc"
    toc: true
    number-sections: false
    code-overflow: wrap
author: "YOUR NAME"
date: "`r Sys.Date()`"
---

### Report Instructions

Graduate students and undergraduates working on independent projects need to provide an overview of what they've been working on at the end of the semester / summer. For undergraduates or anyone else working primarily on lab maintenance, general support, or other broader tasks, your only tasks is to make sure any protocols used are up-to-date and descriptive. This will help me keep the lab running over time.

Reports should be generated with this quarto file, and need to be independent (no additional files needed to render the document). The styling/aesthetics are up to you.

This report should cover work across any and all projects (you don't need a separate report for each one). The report should include:

- A short introduction to each project\
- The relevant hypotheses\
- Detailed methods (or progress on method development)\
- An overview of data collected\
- A section on main conclusions, troubleshooting, and next steps

Once completed (and the draft is approved by Matt), these reports should be rendered and both the qmd file and the PDF output posted to the 'term-reports' channel in slack - make sure they're written so that anyone in the lab can follow along!

A couple small (but important) requirements:

- All code chunks should use `echo = T`.
- Save the file / PDF output as YYYY_LASTNAME_TERM.pdf
  - YYYY = Year in 4-digit format (e.g. 2025)
  - LASTNAME = your last name in upper case (e.g. SASAKI)
  - TERM = Either "FALL", "SPRING", or "SUMMER".
  - Combined, the PDF file name should be something like "2025_SASAKI_FALL.pdf"

The instructions above can be removed before submitting the final report. The sections below can be used for the report (just replace the text under each heading).

------------------------------------------------------------------------

```{r message = F, warning = F, echo=T}

library(tidyverse)
```

### Component Projects

Provide a brief introduction to each projects covered in the report, along with the main hypothesis you're testing (1-2 paragraphs total). It would be useful to include a brief description of the scope of each project included in the report (e.g. PhD chapter, honors thesis, standard lab project).

## Data Sources

Make sure this section is clearly organized. This will typically include at least two sections detailing the relevant project data is stored (i.e. which GitHub repos are the project data stored in?), and the specific data files are you going to use in the report. Whenever possible, data should be read in from GitHub (repos should be public, not private - let Matt know if there's any issues there). Where other data is used, make sure the source is clearly described. Note: you may need a combination of the raw data and the compiled data (i.e. the files in the Output directory).

The example below outlines how this section should be structured:

**Repo 1** - [**starved_ctmax**](https://github.com/ZoopEcoEvo/starved_ctmax/)

This repo is for the project on XYZ. Specific data sets included in this report are:

```{r}

example_data = read.csv("https://raw.githubusercontent.com/ZoopEcoEvo/starved_ctmax/refs/heads/main/Output/Data/starv_data.csv")

```

## Methods

Provide **detailed** methods. Ideally you would provide links to protocols in the lab protocol repository (this is your chance to make sure your protocols are uploaded/updated), along with a detailed outline of the methods used (think manuscript style methods section).

## Data Summary

Clearly describe the data you've collected. Be sure to include basic elements (sample sizes, traits measured, summary stats, etc.) along with description of results or patterns.

When included, figures should be generated using the data read in from the GitHub repo (i.e. write the code to make the figures here in the report). Remember, these term report .qmd files should be **independent** - you should be able to send your .qmd file to someone, and they should be able to render the document with no additional files.

```{r fig.width=8, fig.height=5}

ggplot(example_data, aes(x = rep_day, y = ct_max, colour = diet)) + 
  geom_point() + 
  labs(x = "Day", 
       y = "CTmax (°C)") + 
  theme_bw(base_size = 24) + 
  theme(panel.grid = element_blank())

```

## Conclusions & Troubleshooting

What's the main conclusion you're taking away from the data you've collected. Focus on particular on how findings relate to your hypotheses. Interesting or unexpected observations are also good to discuss here!

## Next Steps

Essentially, what are you going to work on next to advance the project? Focus on the next term, but you can also mention long-term plans.
